Design and Implementation of Probabilistic Programming Language Anglican

### 2017-02-23 paper/paper.tex
ack to Hongseok's funding.
minor (mostly cosmetic) edits

\newtoggle{full}
\input{toggle}

@@

\begin{abstract}
    Anglican is a probabilistic programming system designed to interoperate
    with Clojure and other JVM languages. We introduce the
    programming language Anglican, outline our design choices,
    and discuss in depth the implementation of the Anglican language
    and runtime, including macro-based compilation, extended
    CPS-based evaluation model, and functional representations for
    probabilistic paradigms, such as a distribution,
    a random process, and an inference algorithm.

    We show that a probabilistic functional language can be
    implemented efficiently and integrated tightly with a
    conventional functional language with only moderate
    computational overhead. We also demonstrate how advanced
    probabilistic modeling concepts are mapped naturally
    to the functional foundation.
\end{abstract}

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model. However,
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.

@@

An Anglican program, or \textit{query}, is compiled into a
Clojure function. When inference is performed with a provided
algorithm, this produces a sequence of samples. Anglican shares
a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican
can look like the following code:
%\begin{minipage}{\linewidth}
\begin{lstlisting}[style=default]
(defquery model "model selection" data
  (let [;;; Guess a distribution.
        dist (sample (categorical
                       [[normal 0.5]
                        [gamma 0.5]]))
        a (sample (gamma 1 1))
        b (sample (gamma 1 1))
        d (dist a b)]

@@

%Operational semantics of Anglican queries is different from that
%of Clojure code, therefore queries must be called through
%inference algorithms, rather than `directly'. 
Operational semantics of an Anglican query is different from that
of Clojure code, and is determined by an inference algorithm. Thus,
Anglican queries must be called through these inference algorithms, rather
rather than `directly'. For this purpose,
\iftoggle{full}{the
\texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
function \texttt{infer} using Clojure's multimethod mechanism.
The function accepts and runs an Anglican query, 
and returns a lazy sequence of weighted samples from the
distribution defined by the query. Providing an implementation of this function
is a responsibilit of an inference algorithm, which should also
override the polymorphic function \texttt{checkpoint} (defined
as a multimethod) so as to handle \texttt{sample} and
\texttt{observe} in an algorithm-specific manner and
to construct an appropriate result on the
termination of a probabilistic program.

@@

Anglican programs define probability distributions 
over sequences of values, implicitly by means of program execution.
A good way to understand this is to imagine the following 
interpreter of Anglican programs. Starting from a fixed initial state,
the interpreter runs the deterministic parts of a program 
according to the standard semantics, executes the \texttt{sample} form by
generating a random sample, and treats the $\texttt{observe}$ form by skip.
More importantly, the interpreter keeps a log that records
information about all the \texttt{sample} and \texttt{observe} forms encountered 
during execution. The information recorded for \texttt{sample} is 
a triple $(F,x,\alpha)$ of (i) a primitive probability distribution $F$, 
such as the standard normal, for which we have 
the probability density $p_F$; (ii) a value $x$ sampled from the distribution $F$; 
and (iii) an address $\alpha$ that uniquely and systematically identifies the random 
choice made. The information recorded for \texttt{observe} is a pair $(G,y)$ where $G$
is a primitive probability distribution as $F$ from above and $y$ is an 
observed value. Thus, a log is a sequence of triples $(F,x,\alpha)$ and 
pairs $(G,y)$.

One important property of logs is that they are determined by 
their projections to triples: when two logs project to the same
sequence of triples, they must be the same. This is because
the triples in a log contain all the information about random choices
made during execution and all the non-sample forms in Anglican programs
are deterministic. We define a \textit{trace} $\pmb{x}$ to be a sequence
of triples $(F,x,\alpha)$, and say that $\pmb{x}$ is feasible if
the trace is precisely the triple part of the log of some execution. 
Such a feasible trace uniquely determines the rest of the execution and its log.
In particular, it decides the pair part of the log, namely, 
a sequence of $(G_j, y_j)$ for observed values. We call this sequence \textit{image of $\pmb{x}$} 
and denote it by~$\pmb{y}$.

A (almost-surely terminating) probabilistic program defines a probability distribution 
over finite feasible traces $\pmb{x}$. The distribution has the probability density 
$\pi(\pmb{x}) := \gamma(\pmb{x}) / Z$ where 
\begin{align}
  \gamma(\pmb{x}) 
  &:= 
  \prod_{i=1}^{\left|\pmb{x}\right|}
  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j})
  \quad \mbox{for the image $\pmb{y}$ of $\pmb{x}$}, 
  \label{eqn:p-trace}
\end{align}
and $Z$ is the normalization constant $Z := \int \gamma(\pmb{x})d(\pmb{x})$
of $\gamma$.\footnote{The integral and the density use the default measure obtained by the standard extension of
a $\sigma$-finite measure on finite traces where the $\sigma$-finite measure itself is defined by
a countable sum of Lebesque and counting measures.}

@@

A good way to understand the \texttt{sample} and \texttt{observe} forms
is to consider the imaginary interpreter in Section~\ref{sec:pp-def}.
Under this interpreter, the \texttt{sample} form draws a value $x$ from 
its parameter distribution $F$, and 
generates a unique address $\alpha$ for this random draw.
Then, it appends $(F, x, \alpha)$, the record of this random
draw, to the log of the current execution. Finally, the
form returns $x$. The computation of the \texttt{observe} form is
simpler. It just appends its two parameters, distribution $G$
and observed value $v$, to the log of the current execution,
and returns \texttt{nil}.

What we have just described is idealised semantics 
of \texttt{sample} and \texttt{observe}. The actual computation 
involved in handling \texttt{sample} and \texttt{observe} 
depends on the inference algorithm. Different inference algorithms may treat
\texttt{sample} and \texttt{observe} differently as long as they
compute the same distribution on traces (i.e. sequences of sampled values), 
namely, the one in Equation~\ref{eqn:p-trace}.
Formal semantics for an idealized version
of Anglican or other higher-order probabilistic programming languages are introduced 
in \cite{SYH+16,BorgstromLGS16}; in general though,
specifying semantics of higher-order probabilistic programming
languages is an open problem.

%Their operational semantics can be defined in terms of
%importance sampling(Equation~\ref{eqn:p-trace}): 
%\begin{itemize}
%    \item \texttt{sample} returns a value drawn from
%        the distribution.
%    \item \texttt{observe} multiplies the unnormalized
%        probability of the program trace by the probability of
%        drawing the value from the distribution.
%\end{itemize}
%However, the actual computation involved in handling of
%\texttt{sample} and \texttt{observe} depends on the inference
%algorithm. Different inference algorithms may treat
%\texttt{sample} and \texttt{observe} differently as long as they
%compute the same distribution of program results.
%Denotational and operational semantics for an idealized version
%of Anglican is introduced in \cite{SYH+16}; in general though,
%specifying semantics of higher-order probabilistic programming
%languages is an open problem.

@@

The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithms. \texttt{\$state}
is a Clojure hash map:
\begin{lstlisting}[style=default]
(def initial-state
  "initial program state"
  {:log-weight 0.0, ;; map :log-weight to 0.0
   :result nil,       ;; map :result to nil
   ::store nil,       ;; map :store to nil
   ::mem {},          ;; map ::mem to the empty map 
   ... })
\end{lstlisting}
which records inference-relevant information under various keys
such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
The full list of map entries depends on the inference
algorithm. Except for transformation of
the \texttt{mem} form (which converts a function to one with memoization), 
CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by  returning a \textbf{checkpoint} instead of
invoking a continuation in a CPS-transformed program.  
Checkpoints are values of the
corresponding record types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe})
in Clojure. The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
\begin{lstlisting}[style=default]
=> (cps-of-expression '(sample dist) 'cont)
(->sample dist cont $state)
=> (cps-of-expression '(observe dist v) 'cont)
(->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.

@@

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument  and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program. Hence, \texttt{mem}'s
memory is stored as a nested dictionary in the program state
introduced during CPS transformation (function \texttt{mem-cps}).  
Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same argument. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn [] (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply foo
        ;; memoize result in state
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state P)))))
\end{lstlisting}

@@

When an Anglican function is transformed into a Clojure
function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary
parameters are added to the beginning of the parameter list ---
continuation and state.  Correspondingly, when a function
\textit{call} is transformed\iftoggle{full}{ (by
\texttt{cps-of-application} or \texttt{cps-of-apply})}{}, the
current continuation and the state are passed to the called
function. Anglican can also call Clojure functions; however,
Clojure functions do not expect these auxiliary parameters.
To allow the mixing of Anglican (CPS-transformed) and Clojure
function calls in Anglican code, the Anglican compiler must CPS-transform
expressions selectively~\cite{Nielsen01}. It must be
able to recognize `primitive' (that is, implemented in Clojure
rather than in Anglican) functions, and invoke
those functions in a direct, not CPS, style.

>>>>

\newtoggle{full}
\input{toggle-ifl}

@@

\begin{abstract}
    Anglican is a probabilistic programming system designed to interoperate
    with Clojure and other JVM languages. We introduce the
    programming language Anglican, outline our design choices,
    and discuss in depth the implementation of the Anglican language
    and runtime, including macro-based compilation, extended
    CPS-based evaluation model, and functional representations for
    probabilistic paradigms, such as a distribution,
    a random process, and an inference algorithm.
    We show that a probabilistic functional language can be
    implemented efficiently and integrated tightly with a
    conventional functional language with only moderate
    computational overhead. We also demonstrate how advanced
    probabilistic modeling concepts are mapped naturally
    to the functional foundation.
\end{abstract}

@@

Probabilistic programming systems \cite{GMR+08,MSP14,WVM14,GS15}
represent generative models as programs written in a specialized
language that provides syntax for the definition and conditioning of
random variables. The code for such models is generally concise,
modular, and easy to modify or extend. Typically inference can be
performed for any probabilistic program using one or more generic
inference techniques provided by the system backend, such as
Metropolis-Hastings \cite{WSG11,MSP14,YHG14}, Hamiltonian Monte Carlo
\cite{SDT14}, expectation propagation \cite{MWG+10}, and extensions
of Sequential Monte Carlo \cite{WVM14,MYM+15,PWD+14} methods.
These generic techniques may be less statistically efficient
than techniques tailored to a specific model. However,
probabilistic programming facilitates simpler implementations of
models making the inference inherently faster.
\jw{How exactly does probabilistic programming make inference inherently faster?}

@@

An Anglican program, or \textit{query}, is compiled into a
Clojure function. When inference is performed with a provided
algorithm, this produces a sequence of samples. Anglican shares
a common syntax with Clojure; Clojure functions can be called
from Anglican code and vice versa. A simple program in Anglican
can look like the following code:
%\begin{minipage}{\linewidth}
\begin{lstlisting}[style=default]
  (defquery model "model selection" [data]
    (let [;;; Guess a distribution.
          dist (sample (categorical
                         [[normal 0.5]
                          [gamma 0.5]]))
          a (sample (gamma 1 1))
          b (sample (gamma 1 1))
          d (dist a b)]

@@

%Operational semantics of Anglican queries is different from that
%of Clojure code, therefore queries must be called through
%inference algorithms, rather than `directly'. 
The operational semantics of an Anglican query are different from those
of Clojure code, and are determined by an inference algorithm. Thus,
Anglican queries must be called through these inference algorithms, rather
rather than `directly'. For this purpose,
\iftoggle{full}{the
\texttt{anglican.inference} namespace}{Anglican} declares the (ad-hoc) polymorphic
function \texttt{infer} using Clojure's multimethod mechanism.
This function accepts and runs an Anglican query, 
and returns a lazy sequence of weighted samples from the
distribution defined by the query. Providing an implementation of this function
is a responsibility of the inference algorithm, which should also
override the polymorphic function \texttt{checkpoint} (defined
as a multimethod) so as to handle \texttt{sample} and
\texttt{observe} in an algorithm-specific manner and
to construct an appropriate result on the
termination of a probabilistic program.

@@

\hy{Need to be revised.}

Programs in Anglican define probability distributions 
over sequences of sampled values through their executions.

Imagine an interpreter of Anglican that runs the deterministic parts of a Anglican program 
according to its standard semantics, executes its \texttt{sample} form by
generating a random sample, and treats its $\texttt{observe}$ form by skip.
In addition to this execution, the interpreter keeps a log that records
information about all the \texttt{sample} and \texttt{observe} forms encountered 
during execution. The information kept for \texttt{sample} is a triple $(F,x,\alpha)$ of
the following data:
\begin{itemize}
        \item $F$ is one of the built-in primitive probability distributions, 
                such as the standard normal, for which we have
                the probability density or probability mass function $p_F$.
        \item $x$ is a value sampled from the distribution $F$.
        \item $\alpha$ is an address that uniquely and systematically identifies the random 
                choice made; this address is exploited in inference algorithms, as
                explained later.
\end{itemize}
The information kept for \texttt{observe} is a pair $(G,y)$ where $G$
is a primitive probability distribution as $F$ above and $y$ is an observed value.
Thus, a log is just a sequence of triples $(F,x,\alpha)$ and pairs $(G,y)$.

We call the triple part of a log \textit{trace}. This part records all the random 
choices of an execution. Thus, if the trace of a log is fixed, the rest of the log 
is decided determistically by the program. A trace induces
a sequence of pairs $(G_j, y_j)$ of distributions and values of
\textit{observed} random variables. We call this sequence an \textit{image} and
denote it by $\pmb{y}$. For notational simplicity we assume that the program
always generates the same ordered set $\pmb{z}$ of \textit{results} $z_k$.

The target density $\pi(\vec{x}) := \gamma(\vec{x}) / Z$ of a program
is defined in terms of the product of the probabilities of all random
choices $\pmb{x}$ and the likelihood of all observations $\pmb{y}$
\begin{align}
  \gamma(\pmb{x}) 
  &:= 
  \prod_{i=1}^{\left|\pmb{x}\right|}
  p_{F_i}(x_i) \prod_{j=1}^{\left|\pmb{y}\right|}p_{G_j}(y_{j}).
  \label{eqn:p-trace}
\end{align}
The objective of inference in probabilistic program $\mathcal{P}$
is to characterize the distribution of $\pmb{z}$.

@@

\texttt{sample} and \texttt{observe} forms correspond to
handling $(F, \alpha)$ and $(G, y)$ correspondingly in the
definition of probabilistic program (Section~\ref{sec:pp-def}).
Their operational semantics can be defined in terms of
importance sampling(Equation~\ref{eqn:p-trace}): 
\begin{itemize}
    \item \texttt{sample} returns a value drawn from
        the distribution.
    \item \texttt{observe} multiplies the unnormalized
        probability of the program trace by the probability of
        drawing the value from the distribution.
\end{itemize}
However, the actual computation involved in handling of
\texttt{sample} and \texttt{observe} depends on the inference
algorithm. Different inference algorithms may treat
\texttt{sample} and \texttt{observe} differently as long as they
compute the same distribution of program results.
Denotational and operational semantics for an idealized version
of Anglican is introduced in \cite{SYH+16}; in general though,
specifying semantics of higher-order probabilistic programming
languages is an open problem.

@@

The state (\texttt{\$state}) is threaded through the computation
and contains data used by inference algorithms. \texttt{\$state}
is a Clojure hash map:
\begin{lstlisting}[style=default]
  (def initial-state
    "initial program state"
    {:log-weight 0.0,
     :result nil,
     ::mem {},
     ::store nil,
     ... })
\end{lstlisting}
which records inference-relevant information under various keys
such as $\texttt{:log-weight}$ and $\texttt{::mem}$.
The full list of map entries depends on the inference
algorithm. Except for transformation of
the \texttt{mem} form (which converts a function to one with memoization), 
CPS transformation routines
are not aware of contents of \texttt{\$state}, do not access or
modify it directly, but rather just thread the state unmodified
through the computation. Algorithm-specific handlers of
checkpoints corresponding to the probabilistic forms
(\texttt{sample}, \texttt{observe}) modify the
state and reinject a new state into the computation.

@@

The purpose of probabilistic forms \texttt{sample} and
\texttt{observe} is to interrupt deterministic computation and
transfer control to the inference algorithm.  Practically, this
is achieved by  returning a \textbf{checkpoint} instead of
invoking a continuation in a CPS-transformed program.  In
Clojure, checkpoints are represented by values of the
corresponding record types
(\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}sample} or
\texttt{anglican.{\linebreak[0]}trap.{\linebreak[0]}observe}).
The types contain fields specific to each form, as well as the
continuation; calling the continuation resumes the computation.
\begin{lstlisting}[style=default]
  => (cps-of-expression '(sample dist) 'cont)
  (->sample dist cont $state)
  => (cps-of-expression '(observe dist v) 'cont)
  (->observe dist v cont $state)
\end{lstlisting}
Here \texttt{->sample} and \texttt{->observe} in the
CPS-transformed expressions are constructors for Clojure
records, and they take values of their fields as arguments.

@@

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program. Hence, \texttt{mem}'s
memory is stored as a nested dictionary in the program state
introduced during CPS transformation (function \texttt{mem-cps}).  
Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
  => (mem-cps '(foo))
  (let [M (gensym "M")]
    (fn mem23623 [C $state & P]
      (if (in-mem? $state M P)
        ;; previously memoized result
        (fn [] (C (get-mem $state M P) $state))
        ;; new computation
        (clojure.core/apply foo
          ;; memoize result in state
          (fn [V $state]
            (fn [] (C V (set-mem $state M P V))))
          $state P)))))
\end{lstlisting}

@@

When an Anglican function is transformed into a Clojure
function\iftoggle{full}{ by \texttt{fn-cps}}{}, two auxiliary
parameters are added to the beginning of the parameter list ---
continuation and state.  Correspondingly, when a function
\textit{call} is transformed\iftoggle{full}{ (by
\texttt{cps-of-application} or \texttt{cps-of-apply})}{}, the
current continuation and the state are passed to the called
function. Anglican can also call Clojure functions; however,
Clojure functions do not expect these auxiliary parameters.
To allow the mixing of Anglican (CPS-transformed) and Clojure
function calls in Anglican code, the Anglican compiler must be
able to recognize `primitive' (that is, implemented in Clojure
rather than in Anglican) functions.
